Summer 2025 ToDo

### 2025-10-29 Relations/Paper/Introduction.tex
Renamed implications section to hierarchy. Updated hierarchy section.
Intro update

This broader goal also drives the main design choices in our approach.
Since rewriting theory is concerned with a fine-grained analysis of computation,
the most natural vehicle for formalizing it is type theory based on the
Curry--Howard isomorphism, which Agda implements.  Proofs in this language
are automatically effective, and implicitly carry the code implementing the
transformations from the hypotheses to the conclusion.  For example,
a proof that a given TRS is terminating automatically renders a function that
computes a normal form of any term in its domain.

@@

Now, suppose $T$ admits a convergent presentation.
That is, suppose that the equivalence relation on $S^*$ coincides
with conversion generated by a confluent and terminating rewriting system
${\cdot}{\to_R}{\cdot} : S^* \to S^* \to \bset$.
Then one can replace $S^*/{\approx_T}$ by the type of $R$-normal forms.
(In the semiring example, these would be appropriately sorted polynomial expressions.)
This can be defined as a pure dependent type
\[ NF(R) \quad := \quad (\Sigma n : S^*) (\Pi x : S^*) \lnot (n \to_R x) \]
This type encodes the quotient in the sense that the type of functions $S^*/{\approx_T} \to A$ is isomorphic to plain functions from $NF(R)$ to $A$.
Indeed, the elimination rule for the quotient type requires every equation in $T$
to be validated in order to define a function $f : S^*/{\approx_T} \to A$.
Any such function trivially specializes to $NF(R)$.  In the other direction,
a plain function from $NF(R)$ to $A$
can be extended to all of $S^*$ by induction on normalizing reduction sequences.
Confluence ensures that $T$-equal elements are always mapped to the same value.

@@

\begin{itemize}
  \item No use of function extensionality;
  \item No use of univalence, uniqueness of identity proofs, axiom K, or any other
  assumptions related to equality;
  \item Minimize the use of classical logic, and make decidability hypotheses explicit in every place where classical logic could not be avoided;
  \item Stay as faithful as possible to the spirit of type theory based on the Curry--Howard isomorphism, so that in every explicit application, our proofs would compute.
\end{itemize}

@@

Since Strong Normalization is a central concept in ARS theory,
special attention in our work was given to various incarnations of the concept of well-foundedness in the constructive context.
(Strong normalization is well-foundedness of the converse relation.)
The standard constructive definition does not even allow one to show that SN implies WN, as we discuss in Section \ref{sec:Well-foundedness} and formalize in \texttt{WFCounters.agda}.
  We therefore looked at a number of variations of this notion, their classical counterparts all being logically equivalent. %This gave rise to a rich set of concepts, see Figure \ref{f:wf}.

@@

The structure of the paper roughly corresponds with the progression of the Agda code.
% Agda files.
% (with the exception of section \ref{sec:Well-foundedness}, which is put at the end)
In the next section, we lay out the main properties of relations studied in ARS theory,
setting the ground for what follows.  In Section \ref{sec:Formalization}, we
outline our formalization of the ARS chapter from \cite{Terese}.
Here we also discuss how this effort suggested new ARS properties that helped us overcome some obstacles we encountered along the way.  These properties fit naturally within the rewriting paradigm.  Thus in Section \ref{sec:Implications}, we dedicate some effort to investigating them further.  The result is a broader ontology of ARS concepts, revealing interrelationships between conditions for completeness.  In Section \ref{sec:Well-foundedness} we focus on what it means for a relation to be well-founded, identifying the classical principles needed to transition between different formulations.
We conclude with a brief summary and suggestions for future work.

>>>>

This broader goal also drives the main design choices in our approach.
Since rewriting theory is concerned with a fine-grained analysis of computation,
the most natural vehicle for formalizing it is type theory based on the
Curry--Howard isomorphism, which Agda implements.  Proofs in this language
are automatically effective, and implicitly carry the code that implements
a transformation from the hypotheses to the conclusion of the claim.  For example,
a type-checked proof that a given TRS is confluent automatically renders a function that
computes a common reduct between any two reducts of a given term.

@@

Now, suppose $T$ admits a convergent presentation.
That is, suppose that the equivalence relation on $S^*$ coincides
with conversion generated by a confluent and terminating rewriting system
$\to_R : S^* \to S^* \to \bset$.
Then one can replace $S^*/{\approx_T}$ by the type of $R$-normal forms.
(In the semiring example, these would be polynomial expressions sorted according to 
an appropriate ordering on monomials.)
This can be defined as a pure dependent type
\[ NF(R) \quad := \quad (\Sigma n : S^*) (\Pi x : S^*) \lnot (n \to_R x) \]
This type encodes the quotient in the sense that the type of functions $S^*/{\approx_T} \to A$ is isomorphic to plain functions from $NF(R)$ to $A$.
Indeed, the elimination rule for the quotient type requires every equation in $T$
to be validated in order to define a function $f : S^*/{\approx_T} \to A$.
Any such function trivially specializes to $NF(R)$.  In the other direction,
a plain function from $NF(R)$ to $A$
can be extended to all of $S^*$ by induction on normalizing reduction sequences.
Confluence ensures that $T$-equal elements are always mapped to the same value.

@@

\begin{itemize}
  \item No use of function extensionality;
  \item No use of univalence, uniqueness of identity proofs, axiom K, or any other
  assumptions related to equality;
  \item Minimize the use of classical logic, and make decidability hypotheses explicit in every place where classical logic could not be avoided;
  \item Stay as faithful as possible to the spirit of type theory based on the Curry--Howard isomorphism, so that in every concrete application, our proofs would compute.
\end{itemize}

@@

Recall that a relation is strongly normalizing iff its converse is well-founded. \cite[Def.1.1.13]{Terese}.
Since normalization is a central concept in ARS theory,
we pay special attention to the concept of well-foundedness.
The standard constructive definition does not even allow one to show that SN implies WN, 
as we discuss in Section \ref{sec:Well-foundedness} and formalize in \texttt{WFCounters.agda}.
We therefore looked at a number of variations of this notion in the constructive context, 
their classical counterparts all being logically equivalent. %This gave rise to a rich set of concepts, see Figure \ref{f:wf}.

@@

The structure of the paper roughly corresponds with the progression of the Agda code.
% Agda files.
% (with the exception of section \ref{sec:Well-foundedness}, which is put at the end)
In the next section, we lay out the main properties of relations studied in ARS theory,
setting the ground for what follows.  In Section \ref{sec:Formalization}, we
outline our formalization of the ARS chapter from \cite{Terese}.
Here we also discuss how this effort suggested new ARS properties that helped us overcome some obstacles we encountered along the way.  These properties fit naturally within the rewriting paradigm.  Thus in Section \ref{sec:Hierarchy}, we dedicate some effort to investigating them further.  The result is a broader ontology of ARS concepts, revealing interrelationships between conditions for completeness.  In Section \ref{sec:Well-foundedness} we focus on what it means for a relation to be well-founded, identifying the classical principles needed to transition between different formulations.
We conclude with a brief summary and suggestions for future work.
